Visualization and analysis of probability distributions of large temporal data

### 2022-02-07 thesis.Rmd
Figure 2.4 Catherine

# Copyright notice {- #ch:copyright}

@@

# Abstract {- #ch:abstract}

The research was motivated by the desire to understand the Australian smart meter data, which was collected half-hourly for two years at the household level. With temporal data available at ever finer scales,
exploration may be required to be carried out across both finer and coarser scales to draw useful inferences about the underlying process.
Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task daunting.

@@

This work investigates how time may be dissected, resulting in alternative data segmentation and, as a result, different visualizations that can aid in the identification of underlying patterns. The first contribution (Chapter \@ref(ch:gravitas)) describes classes of time deconstructions using linear and cyclic time granularities. It provides tools to compute possible cyclic granularities from an ordered (usually temporal) index and also a framework to systematically explore the distribution of a univariate variable conditional on two cyclic time granularities by defining "harmony". "harmony" denotes pairs of granularities that could be effectively analyzed together and reduces the search from all possible options. This approach is still overwhelming for human consumption due of the vast amount of harmonies remaining. The second contribution (Chapter \@ref(ch:hakear)) refines the search for informative granularities by identifying those for which the differences between the displayed distributions are greatest and also rating them in order of importance of capturing maximum variation. The third contribution (Chapter \@ref(ch:gracsr)) builds upon the first two to provide methods for exploring heterogeneities in repetitive behavior for many time series and over multiple granularities. It accomplishes this by providing a way to cluster time series based on probability distributions across informative cyclic granularities. Although we were motivated by the smart meter example, the problem and the solutions we propose are practically relevant to any temporal data observed more than once per year.

@@

# Declaration {-#ch:declare}

I hereby declare that this thesis contains no material which has been accepted for the award of any other degree or diploma at any university or equivalent institution and that, to the best of my knowledge and belief, this thesis contains no material previously published or written by another person, except where due reference is made in the text of the thesis.

Three publications are included in this thesis, one of which has been accepted for publication in the Journal of Computational and Graphical Statistics (@Gupta2021-gravitas). Two others have not yet been submitted. "Exploring probability distributions across cyclic temporal granularities" is the central theme of the thesis. All of the papers in the thesis were conceptualized, developed, and written by myself, the student, while working in the Department of Econometrics and Business Statistics under the supervision of Professor Rob J Hyndman and Professor Dianne Cook. When co-authors are listed, it indicates that the work was produced through active collaboration between researchers and that their contributions to team-based research have been recognized. The following table details the work, including
my and my fellow co-authors contributions:

\begin{table}
\centering\footnotesize\tabcolsep=0.12cm
\begin{tabular}{|p{1cm}|p{2cm}|p{1.5cm}|p{3.5cm}|p{3.5cm}|p{1.5cm}|}
\hline
\RaggedRight\textbf{Thesis Chapter}  & \RaggedRight\textbf{Publication Title}  & \RaggedRight\textbf{Status (published, in press, accepted or returned for revision)}  & \RaggedRight\textbf{Nature and} {\%} \RaggedRight\textbf{of student contribution} & \RaggedRight\textbf{Co-author name(s) Nature and} {\%} \RaggedRight\textbf{of Co-author’s contribution} & \RaggedRight\textbf{Co-author(s), Monash student Y/N} \\ \hline
2 & Visualizing probability distributions across bivariate cyclic temporal granularities & Published & 70\%. Concept and developing software and writing first draft & (1) Rob J Hyndman, input into manuscript 15\% (2) Dianne Cook, Concept and input into manuscript 15\% & \multicolumn{1}{c|}{N} \\
\hline
3 & Detecting distributional differences between temporal granularities
for exploratory time series analysis & 80\%. Concept and developing software and writing first draft & (1) Rob J Hyndman, Concept and input into manuscript 10\% (2) Dianne Cook, Concept and input into manuscript 10\%  & \multicolumn{1}{c|}{N} \\
\hline
4 & Clustering time series based on probability distributions across temporal granularities & To be submitted & 70\%. Concept and developing software and writing first draft & (1) Dianne Cook, Concept and input into manuscript 20\% (2) Rob J Hyndman, concept and input into manuscript 10\% & \multicolumn{1}{c|}{N} \\
\hline
\end{tabular}
\end{table}

\FloatBarrier

I have not renumbered sections of submitted or published papers in order to generate a consistent presentation within the thesis.

@@

```{r signature, out.width = "20%", fig.align = "left", eval = file.exists("img/signature.png")}
knitr::include_graphics("img/signature.png")
```

\textbf{Date:} `r Sys.Date()`

@@

Rob and Di are phenomenal leaders in their respective fields who lead by example. I am inspired by their creativity, wisdom, discipline, and dedication to contribute to the society through their research. Thank you for constantly pushing me to improve as a programmer and researcher and regularly sharing good practices for conducting research. Looking back, I am grateful for how my thoughts and work in statistical computing, graphics, and data analytics, in general, have evolved over the years. I still have a lot to learn, but I am a much more self-reliant and independent researcher than when I started. Di has been instrumental in exposing me to the potential benefits of effective data visualization. As a female researcher, I am encouraged by her willingness to pursue unconventional avenues and have frequently noticed her make conscious and proactive choices to question existing stereotypes and biases. Thank you, Di, for being a fantastic female role model.

@@

I want to thank the Department of Econometrics and Business Statistics, Monash Business School, ARC Centre of Excellence for Mathematical and Statistical Frontiers (ACEMS), and Data61 CSIRO for their financial support. Thanks to the Department, I attended a few conferences in person (before COVID hit) and had a great time networking and being inspired by many people. A few highlights are UseR! 2018 (Brisbane), Young Statistician's Conference 2019 (Canberra), WOMBATS 2019 (Melbourne), and RopenSci 2019 (Sydney). Thank you to Gael Martin, Farshid Wahid, Klaus Ackermann, and Didier Nibbering from my PhD committee for providing such a supportive environment.

 <!-- - PhD friends and R community -->
Thank you, Stephanie and Nicholas Tierney, for being so warm, kind and welcoming when I began my PhD. Being a part of such an inclusive environment meant a lot to me as an international student. Thank you, Nicholas Spyrison and Mitchell O'Hara-Wild, for the numerous informative discussions, brainstorming sessions, as well as for generously sharing your expertise and experience in software development. Thank you to each of you, NUMBATs, and the broader R community for directly and indirectly helping me countless times. You never fail to inspire me.

@@

PhD in Covid has been harder, and I would not have been able to make it through this if not for the emotional support of my friends and family. Thank you Puwasala, Sium, and Ian, for being so kind and acting as a pillar of emotional support, especially during the last leg of my journey. Thank you Tushar, Samarpita and Nairita for always motivating me and believing in me when I did not believe in myself. Thank you to each one of you for being available, listening to me, and putting things in perspective when I lost sight. Thanks to my housemates Anjali, Surbhi, and Dulaji for the fun company and conversations that kept me sane in the covid lockdowns.

@@

Finally, a big thank you to my family for always being supportive of my choices. Thanks to my mum (Nupur Gupta) and dad (Arun Prasad Gupta), from whom I learnt that no matter where you start from, if you persevere and are sincere in your efforts, you can sail through any difficult situation. They are my constant cheerleaders for all little and big endeavors. Thanks to my brother (Avijit) for always having my back and inspiring me to dream bigger. Thank you, Juhi (sister-in-law), for reminding me of the value of organization in all aspects of life and the importance of prioritizing my physical health from time to time. Thank you all for your part in supporting me in this journey. I love you.

<!-- add people who are helping you with proofreading -->
<!-- Add Peter -->
<!-- Although the journey is coming to an end, you wish to take these life learnings forward -->
<!-- add Earo and Stuart that you followed their model of workflow and templates, using the power of reproducible and accessible  research -->

@@

# Preface {-#ch:preface}

Chapter \@ref(ch:gravitas) has been accepted by the *Journal of Computational and Graphical Statistics*. It has won the ACEMS Business Analytics Prize in 2020. The accompanying R package **gravitas** is on CRAN. Chapter \@ref(ch:hakear) and Chapter \@ref(ch:gracsr) are yet to be submitted.

@@

This thesis is written in R Markdown [@rmarkdown] with **bookdown** [@bookdown], using **renv** [@renv] to create reproducible environments. The online version of this thesis is hosted at <https://sayani.netlify.app/>, powered by [Netlify](http://netlify.com). All materials (including the data sets and source files) required to reproduce this document can be found at the Github repository <https://github.com/Sayani07/thesis>.

@@

The Smart Grid, Smart City (SGSC) project (2010–2014) available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee) provides half-hourly data of over 13,000 household electricity smart meters distributed unevenly from October 2011 to March 2014 for some parts in Australia. Because of the increasing variety of customers, larger data sets contain more uncertainty about customer behavior. Household behavioral patterns vary significantly due to differences in size, location, and amenities such as solar panels, central heating, and air conditioning. For example, some families use a dryer, while others hang their clothes to dry. This could be reflected in their weekly profile. They may vary on a monthly basis, with some customers using more air conditioners or heaters than others despite having comparable electrical equipment and weather conditions. Some customers are night owls, while others are morning larks, which may show up in their daily profile. Customers' day-off energy consumption varies depending on whether they stay at home or go outside. The motivation for this work comes from the desire to provide methods to better understand heterogeneity in repetitive behavior that can arise across temporal dependencies.

<!-- introducing the overall field-->
With the availability of data at more and more finer time scales, exploration of time series data may be required to be carried out across both finer and coarser scales to draw useful inferences about the underlying process. For example, data collected at an hourly scale could be analyzed using coarser temporal scales such as days, months or quarters. This approach requires deconstructing time in various possible ways. Moreover, it might often be interesting to capture calendar or periodic effects like month-of-year, day-of-week, or hour-of-day. They help us answer questions like if certain levels of those time deconstructions are characterized by unusual/typical values of the observed variable. For example, certain days of the week or months of the year are likely to be characterized by higher values. Visualizations using linear time granularities are often ineffective in conveying repetitive behaviors because multiple observations are squeezed together in that representation. Hence, we define cyclic granularities for exploring repetitive patterns. There could potentially be many, many cyclic granularities to explore. It is essential to navigate through all the options to get a comprehensive view of the repeated patterns. This idea aligns with the notion of EDA (Tukey 1977), which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing.

@@

 <!-- This chapter contains a literature review related to linear granularities. -->

@@

This work is a natural extension of Chapter \@ref(ch:gravitas). Many, many other displays might be built using different granularities, such as day-of-week, day-of-month, weekday/weekend, and so on. However, only a handful of them may be interesting, revealing major patterns. Determining which displays exhibit "significant" distributional differences between cyclic granularity categories and plotting only these would allow for more efficient exploration. Furthermore, a few of the displays in this collection will be more engaging than others. This chapter provides a new distance metric for selecting and ranking the multiple granularities. The statistical significance of potential visual discoveries is aided by selecting a threshold for the proposed numerical distance measure. The distance measure is computed for a single or pairs of cyclic granularities, and it can be compared across different cyclic granularities as well as a collection of time series. This chapter also includes a case study using residential smart meter data from Melbourne to demonstrate how the suggested methodology may be utilized to automatically find temporal granularities with significant distributional differences. The methods are implemented in the open-source R package 'hakear.'

@@

In this chapter, we look at the problem of using clustering to discover patterns in a large number of univariate time series across multiple temporal granularities. Time series clustering research is gaining traction as more data is collected at finer temporal resolution, over longer time periods, and for a larger number of individuals/entities. Many disciplines have noisy, patchy, uneven, and asynchronous time series that make it difficult to search for similarities. We propose a method for overcoming these constraints by calculating distances between time series based on probability distributions at various temporal granularities. Because they are based on probability distributions, these distances are resistant to missing or noisy data and aid in dimension reduction. When fed into a clustering algorithm, the distances can be used to divide large data sets into small pockets of similar repetitive behavior. These subgroups can then be analyzed separately or used as distinct prototype behaviors in classification problems. The proposed method was tested on a group of residential electricity consumers from the Australian smart meter data set to show that it can generate meaningful clusters. This chapter includes a brief review of the literature on traditional time series clustering and, more specifically, clustering residential smart meter data.

@@

## Summary

The thesis is structured as follows. Chapter \@ref(ch:gravitas) provides details of the cyclic granularities, different classes, and computation, and also its usage in exploratory time series analysis through applications. This is implemented in the R package **gravitas**. Chapter \@ref(ch:hakear) provides guidance on how to choose significant cyclic granularities, which are likely to have interesting patterns across its categories. This is available as the R package **hakear**. The chapter @ref (ch: gracsr) provides similarity measures for comparing multiple time series.This is in the developing R package **gracsr**. Chapter \@ref(ch:conclusion) summarizes the software tools developed for the work, and discusses some future plans.

@@

If $G$ groups into $H$, it would imply that any granule $H(i)$ is the union of some granules of $G$, for example, $G(a_1), G(a_2), \dots, G(a_k)$. Condition (2) in Definition \ref{def:periodic} implies that if $H(i + R) \neq \emptyset$, then $H(i + R) = \bigcup (G(a_1 + P), G(a_2 + P), \dots, G(a_k + P))$, resulting in a "periodic" pattern of the composition of $H$ using granules of $G$. In this pattern, each granule of $H$ is shifted by $P$ granules of $G$. $P$ is called the \textbf{Period} (@Bettini2000-qk).

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lll}
\toprule
Circular granularity & Expression & Period \\
\midrule
minute-of-hour                               &
  $C_1 = z \mod 60$                     &
  $P_1 = \phantom{99}60$ \\
minute-of-day                                &
  $C_2 = z \mod 60*24$                  &
  $P_2= `r 60*24`$\\
hour-of-day                                  &
  $C_3 = \lfloor z/60\rfloor\mod 24$    &
  $P_3 = \phantom{99}24$ \\
hour-of-week                                 &
  $C_4 = \lfloor z/60\rfloor\mod 24*7$  &
  $P_4= \phantom{9}`r 24*7`$\\
day-of-week                                  &
  $C_5 = \lfloor z/24*60\rfloor \mod 7$ &
  $P_5= \phantom{999}7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and period $P_i$. Circular granularities can be expressed using modular arithmetic due to their regular mapping.}
\label{tab:definitions}
\end{table}

@@

\noindent In the Mayan calendar (Table \@ref(tab:tab-mayan)), kin-of-tun or kin-of-baktun are examples of multiple-order-up granularities and single-order-up granularities are kin-of-uinal, uinal-of-tun etc.

@@

```{r allFig, fig.pos="!p", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap= "(ref:allFig)", fig.show = 'hold'}
```

(ref:allFig) Distribution of energy consumption displayed as letter value plots, illustrating harmonies and clashes, and how mappings change emphasis: **a** weekday/weekend faceted by quarter-of-year produces a harmony, **b** quarter-of-year faceted by weekday/weekend produces a harmony, **c** month-of-year faceted by quarter-of-year produces a clash, as indicated by the empty sets and white space. Placement within a facet should be done for primary comparisons. For example, arrangement in **a** makes it easier to compare across weekday type (x-axis) within a quarter (facet). It can be seen that in quarter 2, there is more mass occupied the lower tail on the weekends (letter value E corresponding to tail area 1/8) relative to that of the weekdays (letter value D 1/16), which corresponds to more days with lower energy use in this period.

@@

The Australian authors thank the ARC Centre of Excellence for Mathematical and Statistical Frontiers [(ACEMS)](https://acems.org.au/home) for supporting this research. Thanks to [Data61 CSIRO](https://data61.csiro.au/) for partially funding Sayani's research and Dr Peter Toscas for providing useful inputs on improving the analysis of the smart meter application. We would also like to thank Nicholas Spyrison for many useful discussions, sketching figures and feedback on the manuscript. The package `gravitas` was built during the [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). More details about the package can be found at [sayani07.github.io/gravitas](https://sayani07.github.io/gravitas/). The Github repository,  [github.com/Sayani07/paper-gravitas](https://github.com/Sayani07/paper-gravitas), contains all materials required to reproduce this article and the code is also available online in the supplemental materials. This article was created with `knitr` [@knitr], [@R-knitr] and `rmarkdown` [@rmarkdown], [@R-rmarkdown].

## Supplementary Materials {- #supplement-gravitas}

@@

# Detecting distributional differences between temporal granularities for exploratory time series analysis  {#ch:hakear}

 Cyclic temporal granularities, which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. The granularities feed new approaches to exploring time series data. One use is to take pairs of granularities, and make plots of response values across the categories induced by the temporal deconstruction. However, when there are many granularities that can be constructed for a time period, there will also be too many possible displays to decide which might be the more interesting to display. This work proposes a new distance metric to screen and rank the possible granularities, and hence choose the most interesting ones to plot. The distance measure is computed for a single or pairs of cyclic granularities can can be compared across different cyclic granularities and also on a collection of time series. The methods are implemented in the open-source R package `hakear`.

@@

```{r onegran-new, fig.cap="A cyclic granularity can be considered to be a categorical variable, and used to break the data into subsets. Here, side-by-side boxplots overlaid on jittered dotplots explore the distribution of of energy use by a household for two different cyclic granularities: (a) hour-of-day and (b) and month-of-year. Daily peaks occur in morning and evening hours, indicating a working household, where members leave for and return from work.  More volatility of usage in summer months (Jan, Feb) is probably due air conditioner use on just some days ."}
```

>>>>

# Copyright notice {- #ch-copyright}

@@

# Abstract {- #ch-abstract}

The research was motivated by the desire to understand some Australian smart meter data, which was collected half-hourly for two years at the household level. With temporal data available at ever finer scales, exploring periodicity can become overwhelming with so many possible temporal deconstructions to explore.
Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly can make the task daunting.

@@

This work investigates how time may be dissected, resulting in alternative data segmentation and, as a result, different visualizations that can aid in the identification of underlying patterns. The first contribution (Chapter \@ref(ch-gravitas)) describes classes of time deconstructions using linear and cyclic time granularities. It provides tools to compute possible cyclic granularities from an ordered (usually temporal) index and also a framework to systematically explore the distribution of a univariate variable conditional on two cyclic time granularities by defining "harmony". A "harmony" denotes pairs of granularities that could be effectively analyzed together and reduces the search from all possible options. This approach is still overwhelming for human consumption due to the vast number of harmonies remaining. The second contribution (Chapter \@ref(ch-hakear)) refines the search for informative granularities by identifying those for which the differences between the displayed distributions are greatest and also rating them in order of importance of capturing maximum variation. The third contribution (Chapter \@ref(ch-gracsr)) builds upon the first two to provide methods for exploring heterogeneities in repetitive behavior for many time series and over multiple granularities. It accomplishes this by providing a way to cluster time series based on probability distributions across informative cyclic granularities. Although we were motivated by the smart meter example, the problem and the solutions we propose are practically relevant to any temporal data observed more than once per year.

@@

# Declaration {-#ch-declare}

This thesis is an original work of my research and contains no material which has been accepted for the award of any other degree or diploma at any university or equivalent institution and that, to the best of my knowledge and belief, this thesis contains no material previously published or written by another person, except where due reference is made in the text of the thesis.

\vspace{4em}

```{r signature, out.width = "20%", fig.align = "left"}
knitr::include_graphics("img/signature.png")
```

@@

\textbf{Date:} `r as.Date("2021-11-19")`

@@

I am deeply grateful to my supervisors Rob J Hyndman and Dianne Cook, who are phenomenal leaders in their respective fields and lead by example. I am inspired by their creativity, wisdom, discipline, and dedication to contribute to the society through their research. Thank you for constantly pushing me to improve as a programmer and researcher and regularly sharing good practices for conducting research. Looking back, I am grateful for how my thoughts and work in statistical computing, graphics, and data analytics, in general, have evolved over the years. I still have a lot to learn, but I am a much more self-reliant and independent researcher than when I started. Di has been instrumental in exposing me to the potential benefits of effective data visualization. As a female researcher, I am encouraged by her willingness to pursue unconventional avenues and have frequently noticed her make conscious and proactive choices to question existing stereotypes and biases. Thank you, Di, for being a fantastic female role model. I would also like to express my gratitude to my supervisor Dr. Peter Toscas of Data61 CSIRO, for always being available to help and offering feedback on practical elements of analyzing smart meter data at our monthly meetings.

@@

I want to thank the Department of Econometrics and Business Statistics, Monash Business School, ARC Centre of Excellence for Mathematical and Statistical Frontiers (ACEMS), Data61 CSIRO for their financial support. Thanks to the Department, I attended a few conferences in person (before COVID hit) and had a great time networking and being inspired by many people. A few highlights are UseR! 2018 (Brisbane), Young Statistician's Conference 2019 (Canberra), WOMBAT 2019 (Melbourne), and ROpenSci 2019 (Sydney). Thank you to Gael Martin, Farshid Wahid, Klaus Ackermann, and Didier Nibbering from my PhD committee for providing such a supportive environment.

 <!-- - PhD friends and R community -->
Thank you, Puwasala Gamakumara, Stephanie Kobakian and Nicholas Tierney, for being so warm, kind and welcoming when I began my PhD. Being a part of such an inclusive environment meant a lot to me as an international student. Thank you, Nicholas Spyrison and Mitchell O'Hara-Wild, for the numerous informative discussions, brainstorming sessions, as well as for generously sharing your expertise and experience in software development. A special thanks goes to Stephanie, Emi Tanaka, Nicholas Tierney, and Nicholas Spyrison for taking the time out of their busy schedules to proofread parts of my thesis. Thank you to each of you, NUMBATs, and the broader R community for directly and indirectly helping me countless times. You never fail to inspire me.

@@

Doing a PhD in COVID has been hard, and I would not have been able to make it through this if not for the emotional support of my friends and family. Thank you Puwasala, Sium, and Ian, for being so kind and acting as a pillar of emotional support, especially during the last leg of my journey. Thank you Tushar, Samarpita and Nairita for always motivating me and believing in me when I did not believe in myself. Thank you to each one of you for being available, listening to me, and putting things in perspective when I lost sight. Thanks to my housemates Anjali, Surbhi, and Dulaji for the fun company, food and conversations that kept me sane in the COVID lockdowns.

@@

A big thank you to my family for always being supportive of my choices. Thanks to my mum (Nupur Gupta) and dad (Arun Prasad Gupta), from whom I learnt that no matter where you start from, if you persevere and are sincere in your efforts, you can sail through any difficult situation. They are my constant cheerleaders for all little and big endeavors. Thanks to my brother (Avijit) for always having my back and inspiring me to dream bigger. Thank you, Juhi (sister-in-law), for reminding me of the value of organization in all aspects of life and the importance of prioritizing my physical health from time to time.
Lastly, thank you to my one-year-old nephew (Rayan) and niece (Mehr) for being the ultimate stress relievers. Thank you all for being a part of this journey with me. I love you.

<!-- Thank you for making me feel at home away from home, Somdatta (cousin). -->
<!-- add people who are helping you with proofreading -->
<!-- Although the journey is coming to an end, you wish to take these life learnings forward -->
<!-- add Earo and Stuart that you followed their model of workflow and templates, using the power of reproducible and accessible  research -->

Throughout my PhD candidature, I have had the great privilege of working with and being inspired by my advisers and many other people, which resulted in a great lot of learning in both academics and life. Although the journey is coming to a close, I want to carry these life lessons forward, and I would like to conclude with one of my favorites from Rob. "Be kind; remember you are collaborating with your future self."

@@

# Preface {-#ch-preface}

Chapter \@ref(ch-gravitas) has been published online at the [_Journal of Computational and Graphical Statistics_](https://www.tandfonline.com/eprint/NXKCSZHIBFXCJD2NKB68/full?target=10.1080/10618600.2021.1938588). It has won the ACEMS Business Analytics Prize in 2020. The accompanying R package `gravitas` is on CRAN. Chapter \@ref(ch-hakear) and Chapter \@ref(ch-gracsr) are yet to be submitted.

@@

This thesis is written in R Markdown [@rmarkdown] with `bookdown` [@bookdown], using `renv` [@renv] to create reproducible environments. The online version of this thesis is hosted at <https://sayani07.github.io/thesis-SG>. All materials (including the data sets and source files) required to reproduce this document can be found at the Github repository <https://github.com/Sayani07/thesis-SG>.

@@

The Smart Grid, Smart City (SGSC) project (2010–2014) available through the [Department of the Environment and Energy](https://data.gov.au/data/organization/doee) provides half-hourly data of over 13,000 Australian electricity smart meter customers distributed unevenly from October 2011 to March 2014. The wide variety of customers means that there will be large variance in behavior, leading to greater uncertainty in the data. Behavioral patterns vary significantly due to differences in size, location, and amenities such as solar panels, central heating, and air conditioning. For example, some families use a dryer, while others hang their clothes to dry. This could be reflected in their weekly profile. They may vary on a monthly basis, with some customers using more air conditioners or heaters than others despite having comparable electrical equipment and weather conditions. Some customers are night owls, while others are morning larks, which may show up in their daily profile. Customers' day-off energy consumption varies depending on whether they stay at home or go outside.

<!-- introducing the overall field-->
With the availability of data at finer and finer time scales, exploration of time series data may be required to be carried out across both finer and coarser scales to draw useful inferences about the underlying process. To reduce the complexity of time, it is typical to divide it into years, months, weeks, days, and so on in a hierarchical manner [@aigner2011visualization]. These discrete abstractions of time are known as time granularities. Linear time granularities [@Bettini1998-ed], such as hours, days, weeks and months, respect the linear progression of time and are non-repeating. Cyclic temporal granularities representing cyclical repetitions in time (such as hour-of-the-day, work-day/weekend) are effective for analyzing repetitive patterns in time series data.

To acquire a comprehensive view of the repeated patterns, it is necessary to navigate through all of the conceivable cyclic granularities. This approach is consistent with the concept of exploratory data analysis EDA [@Tukey1977-jx], which stresses the utilization of multiple perspectives on data to assist with formulating hypotheses before proceeding to formal inferences or modeling. This, however, is a challenging process since it throws up a myriad of possible hypotheses. Furthermore, the transition from linear to cyclic granularities results in restructured data, with each level of the temporal deconstruction corresponding to multiple values of the observed variable. This motivates the research presented in this thesis, which aims to provide a platform for systematically exploring probability distributions induced by these multiple observations to support the discovery of regular patterns or anomalies, as well as the exploration of clusters of behaviors or the summarization of the behavior.

<!-- There could potentially be many, many ways to analyze the data across cyclic granularities. -->

<!-- For example, data collected at an hourly scale could be analyzed using coarser temporal scales such as days, months or quarters. This approach requires deconstructing time in various possible ways. Moreover, it might often be interesting to capture calendar or periodic effects like month-of-year, day-of-week, or hour-of-day. They help us answer questions like if certain levels of those time deconstructions are characterized by unusual/typical values of the observed variable. For example, certain days of the week or months of the year are likely to be characterized by higher values. Visualizations using linear time granularities are often ineffective in conveying repetitive behaviors because multiple observations are squeezed together in that representation. Hence, we define cyclic granularities for exploring repetitive patterns. There could potentially be many, many cyclic granularities to explore. It is essential to navigate through all the options to get a comprehensive view of the repeated patterns. This idea aligns with the notion of EDA (Tukey 1977), which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. -->

@@

Chapter \@ref(ch-gravitas) describes classes of time deconstructions using linear and cyclic time granularities, which can be used to create data visualizations to explore periodicities, associations, and anomalies. It provides a formal characterization of cyclic granularities and facilitates manipulation of single- and multiple-order-up time granularities through cyclic calendar algebra, as well as providing a recommendation algorithm to check the feasibility of creating plots for any two cyclic granularities. Our proposed method is also applicable to non-temporal hierarchical granularities with an underlying ordered index. The methods are implemented in the open-source R package `gravitas` and are consistent with a tidy workflow [@wickham2016r], with probability distributions examined using the range of graphics available in `ggplot2` [@Wickham2009pk].

@@

Chapter \@ref(ch-hakear) is a natural extension of Chapter \@ref(ch-gravitas). Many displays might be built using cyclic granularities. However, only a handful of them may reveal major patterns of interest. Identifying the displays which exhibit "significant" distributional differences and plotting only these would allow for more efficient exploration. Furthermore, a few of the displays in this collection will be more engaging than others. Chapter \@ref(ch-hakear) provides a new distance metric for selecting and ranking the multiple granularities. The statistical significance of potential visual discoveries is aided by selecting a threshold for the proposed numerical distance measure. The distance measure is computed for a single or pairs of cyclic granularities, and it can be compared across different cyclic granularities as well as a collection of time series. This chapter also includes a case study using residential smart meter data from Melbourne to demonstrate how the suggested methodology may be utilized to automatically find temporal granularities with significant distributional differences. The methods are implemented in the open-source R package `hakear`.

@@

In Chapter \@ref(ch-gracsr), we look at the problem of using clustering to discover patterns in a large number of univariate time series across multiple temporal granularities. Time series clustering research is gaining traction as more data is collected at finer temporal resolution, over longer time periods, and for a larger number of individuals/entities. Many disciplines have noisy, patchy, uneven, and asynchronous time series that make it difficult to search for similarities. We propose a method for overcoming these constraints by calculating distances between time series based on probability distributions at various temporal granularities. Because they are based on probability distributions, these distances are resistant to missing or noisy data and aid in dimension reduction. When fed into a clustering algorithm, the distances can be used to divide large data sets into small pockets of similar repetitive behaviors. These subgroups can then be analyzed separately or used as distinct prototype behaviors in classification problems. The proposed method was tested on a group of residential electricity consumers from the Australian smart meter data set to show that it can generate meaningful clusters. This chapter includes a brief review of the literature on traditional time series clustering and, more specifically, clustering residential smart meter data.

@@

## Thesis structure

The thesis is structured as follows. Chapter \@ref(ch-gravitas) provides details of the cyclic granularities, different classes, and computation, and also its usage in exploratory time series analysis through applications. This is implemented in the R package `gravitas`. Chapter \@ref(ch-hakear) provides guidance on how to choose significant cyclic granularities, which are likely to have interesting patterns across its categories. This is available as the R package `hakear`. Chapter \@ref(ch-gracsr) provides methods to explore heterogeneity in repetitive behavior for multiple time series over multiple cyclic granularities. This is in the developing R package `gracsr`. Chapter \@ref(ch-conclusion) summarizes the software tools developed for the work, and discusses future plans.

@@

If $G$ groups into $H$, it would imply that any granule $H(i)$ is the union of some granules of $G$; for example, $G(a_1), G(a_2), \dots, G(a_k)$. Condition (2) in Definition \ref{def:periodic} implies that if $H(i + R) \neq \emptyset$, then $H(i + R) = \bigcup (G(a_1 + P), G(a_2 + P), \dots, G(a_k + P))$, resulting in a "periodic" pattern of the composition of $H$ using granules of $G$. In this pattern, each granule of $H$ is shifted by $P$ granules of $G$. $P$ is called the \textbf{period} [@Bettini2000-qk].

@@

\begin{table}[ht]
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities, one of which groups periodically into the other with regular mapping and period $P_i$. Circular granularities can be expressed using modular arithmetic due to their regular mapping.}
\begin{center}
\begin{tabular}{lll}
\toprule
Circular granularity & Expression & Period \\
\midrule
minute-of-hour                               &
  $C_1 = z \mod 60$                     &
  $P_1 = \phantom{99}60$ \\
minute-of-day                                &
  $C_2 = z \mod 60*24$                  &
  $P_2= `r 60*24`$\\
hour-of-day                                  &
  $C_3 = \lfloor z/60\rfloor\mod 24$    &
  $P_3 = \phantom{99}24$ \\
hour-of-week                                 &
  $C_4 = \lfloor z/60\rfloor\mod 24*7$  &
  $P_4= \phantom{9}`r 24*7`$\\
day-of-week                                  &
  $C_5 = \lfloor z/24*60\rfloor \mod 7$ &
  $P_5= \phantom{999}7$\\
\bottomrule
\end{tabular}
\end{center}
\label{tab:definitions}
\end{table}

@@

\noindent In the Mayan calendar (\autoref{tab:tab-mayan}), kin-of-tun or kin-of-baktun are examples of multiple-order-up granularities and single-order-up granularities are kin-of-uinal, uinal-of-tun etc.

@@

```{r allFig, fig.pos="!p", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap= "(ref:allFig)", fig.show = 'hold', fig.height=6}
```

(ref:allFig) Distribution of energy consumption displayed on a logarithmic scale as letter value plots, illustrating harmonies and clashes, and how mappings change emphasis: **a** weekday/weekend faceted by quarter-of-year produces a harmony, **b** quarter-of-year faceted by weekday/weekend produces a harmony, **c** month-of-year faceted by quarter-of-year produces a clash, as indicated by the empty sets and white space. Placement within a facet should be done for primary comparisons. For example, arrangement in **a** makes it easier to compare across weekday type (x-axis) within a quarter (facet). It can be seen that in quarter 2, more mass occupied the lower tail on the weekends (letter value E corresponding to tail area 1/8) relative to that of the weekdays (letter value D 1/16), which corresponds to more days with lower energy use in this period.

@@

The Australian authors thank the ARC Centre of Excellence for Mathematical and Statistical Frontiers [(ACEMS)](https://acems.org.au/home) for supporting this research. Thanks to [Data61 CSIRO](https://data61.csiro.au/) for partially funding Sayani's research and Dr. Peter Toscas for providing useful inputs on improving the analysis of the smart meter application. We would also like to thank Nicholas Spyrison for many useful discussions, sketching figures and feedback on the manuscript. The package `gravitas` was built during the [Google Summer of Code, 2019](https://summerofcode.withgoogle.com/archive/). More details about the package can be found at [sayani07.github.io/gravitas](https://sayani07.github.io/gravitas/). The Github repository,  [github.com/Sayani07/paper-gravitas](https://github.com/Sayani07/paper-gravitas), contains all materials required to reproduce this article and the code is also available online in the supplemental materials. This article was created with `knitr` [@knitr; @R-knitr] and `rmarkdown` [@rmarkdown; @R-rmarkdown].

## Supplementary materials {- #supplement-gravitas}

@@

# Detecting distributional differences between temporal granularities for exploratory time series analysis  {#ch-hakear}

 Cyclic temporal granularities, which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. The granularities feed new approaches to exploring time series data. One use is to take pairs of granularities, and make plots of response values across the categories induced by the temporal deconstruction. However, when there are many granularities that can be constructed for a time period, there will also be too many possible displays to decide which might be the more interesting to display. This work proposes a new distance metric to screen and rank the possible granularities, and hence choose the most interesting ones to plot. The distance measure is computed for a single or pairs of cyclic granularities that can be compared across different cyclic granularities and also on a collection of time series. The methods are implemented in the open-source R package [`hakear`](https://github.com/Sayani07/hakear).

@@

```{r onegran-new, fig.cap="A cyclic granularity can be considered to be a categorical variable, and used to break the data into subsets. Here, side-by-side boxplots overlaid on jittered dotplots explore the distribution of of energy use by a household for two different cyclic granularities: (a) hour-of-day and (b) and month-of-year. Daily peaks occur in morning and evening hours, indicating a working household, where members leave for and return from work. More volatility of usage in summer months (Jan, Feb) is probably due to air conditioner use on just some days.", fig.pos="!b"}
```
